---
title: "Project Context"
editor_options: 
  chunk_output_type: console
author: "Havisha Khurana"
date: "today"
date-format: "full"
format: 
  closeread-html:
     cr-style:
      narrative-background-color-overlay: darkslategrey
      narrative-text-color-overlay: "#e2e2e2"
      narrative-font-family: "serif"
      section-background-color: "#fffbf1"
     remove-header-space: true
     embed-resources: true
     theme: journal
     mainfont: "serif"
     backgroundcolor: "#fffbf1"
     highlight: tango
     df-print: paged
     code-overflow: wrap
     code-line-numbers: true
     highlight-style: pygments
execute:
   warning: false
   error: false
   message: false
   echo: false
---


```{r}
# library
library(tidyverse)
library(sf)
library(RColorBrewer)
library(readxl)
library(plotly)
library(tigris)

source("code/writeup_plots.R")
```


```{r}
# Download the Hexagon boundaries at geojson format here: https://team.carto.com/u/andrew/tables/andrew.us_states_hexgrid/public/map.

us_hexagon <- read_sf("data_raw/Other Data/us_states_hexgrid.geojson", quiet = TRUE) %>%
  # reformat names
  mutate(google_name = gsub(" \\(United States\\)", "", google_name))

# read sheet 1 on state revenues and expenditures
# re-named columns
cols <- c("geoid", "total_revenue", "federal", "state", "local","total_spending",
          "current_spending", "capital", "other", "debt", "cash_securities")

# data
us_states_revenue22 <- read_excel("data_raw/Other Data/elsec22_sumtables.xls", 
    sheet = "1", range = "A10:L71", col_names = FALSE) %>%
  janitor::clean_names() %>%
  filter(!is.na(x1)) %>%
  select(-x2) %>%
  rename_with(~cols, everything()) %>%
  mutate(
    geoid = gsub("\\.","",geoid)) %>%
  rowwise() %>%
  mutate(
    # convert to billion
    across(-geoid, ~ifelse(!is.numeric(.x), parse_number(.x)/1000000, .x/1000000)),
    # add state revenue percentage
    per_state = (state/total_revenue)*100
  ) %>%
  ungroup()

# read sheet 19 on enrollment
us_states_enrollment <- read_excel("data_raw/Other Data/elsec22_sumtables.xls", 
    sheet = "19", skip = 2) %>%
  janitor::clean_names() %>%
  filter(!is.na(geographic_area)) %>%
  select(geographic_area, elementary_secondary_enrollment) %>%
  mutate(elementary_secondary_enrollment = parse_number(elementary_secondary_enrollment),
         geographic_area = gsub("\\.","",geographic_area)) %>%
  head(., 52)

us_map_data <- us_hexagon %>%
  left_join(us_states_revenue22, by=c("google_name" = "geoid")) %>%
  left_join(us_states_enrollment, by=c("google_name" = "geographic_area")) %>%
  mutate(
    revenue_pp = total_revenue*1000000000 / elementary_secondary_enrollment,
    expense_pp = total_spending*1000000000 / elementary_secondary_enrollment,
    current_expense_pp = current_spending*1000000000 / elementary_secondary_enrollment,
    account_balance = total_revenue - total_spending,
    plotly_text = paste0("State: ", google_name,"\n",
                         "Total Revenue: ",round(total_revenue, 1),"B\n",
                         "State contribution: ", round(per_state),"%\n",
                         "Total Expense: ",round(total_spending, 1),"B\n",
                         "Revenue pp: $", round(revenue_pp),"\n",
                         "Expense pp: $", round(expense_pp),"\n",
                         "Balace: ", round(account_balance, 1),"B")
  )
```


::::::{.cr-section layout="overlay-center"}

:::{#cr-intro}
```{r}
ggplot(us_map_data) +
  geom_sf(fill = "skyblue", color = "white") +
  #geom_sf_text(aes(label = iso3166_2), size = 2) +
  theme_void() +
  theme(
      plot.background = element_blank(),
       panel.background = element_rect(fill =background_color,color = "transparent")
  )
```
:::

An overview of United State's education funding. @cr-intro

::::::


```{r}
state_map_var_discrete(us_map_data, total_revenue, plotly_text, "Total Revenue",
                       subtitle = "In 2022, US education revenue stood at 878 billion dollars.")
```

::::::{.cr-section layout="sidebar-left"}


:::{#cr-totrev}
```{r}
state_map_var_discrete(us_map_data, total_revenue, plotly_text, "Total Revenue",
                       subtitle = "In 2022, US education revenue stood at 878 billion dollars.", return = "plot")
```
:::

In 2022, **878 Billion Dollars** were allocated to school districts in the US. @cr-totrev

Some of the states with the biggest school districts had the most revenue, like the east-coast states [@cr-totrev]{pan-to="-40%,-5%" scale-by="2"}

The states with the lowest revnue were in the Midwest, although, they tend to have lower enrollments too. [@cr-totrev]{pan-to="10%,10%" scale-by="2"}

:::{#cr-revenuepp}
```{r}
state_map_var_discrete(us_map_data, revenue_pp, plotly_text, "Total Revenue Per Pupil",
                       subtitle = "The nation-wide per-pupil revenue was $18,910.", return = "plot")
```
:::

The per-pupil revenue was **$18,910** @cr-revenuepp

Again, there are two regions with higher revenue per-pupil. Northeast [@cr-revenuepp]{pan-to="-40%,20%" scale-by="1.5"}

And, West coast [@cr-revenuepp]{pan-to="35%,10%" scale-by="2"}

States in the center tend to have lower per-pupil revenue. [@cr-revenuepp]{pan-to="10%,10%" scale-by="2"}

:::{#cr-state}
```{r}
state_map_var_discrete(us_map_data, per_state, plotly_text, "State's Contribution",
                       subtitle = "State sources contributed 43% of the total education revenue.", return = "plot")
```
:::

States contributed to **43%** of total revenue. @cr-state

High-state contributors are mostly on the West. [@cr-state]{pan-to="35%,10%" scale-by="2"}

Though there are notable exceptions like, NM, KS, AR [@cr-state]{pan-to="15%,10%" scale-by="2"} 

And, VT [@cr-state]{pan-to="-45%,30%" scale-by="2"}

::: {#cr-spending}
```{r}
state_map_var_discrete(us_map_data, total_spending, plotly_text, "Total Spending",
                       subtitle = "In 2022, US education revenue stood at 857 billion dollars.", return ="plot")
```
:::

Spending has the same geographic trends as revenue generation. @cr-spending

::: {#cr-spendingpp}
```{r}
state_map_var_discrete(us_map_data, expense_pp, plotly_text, "Spending Per Pupil",
                       subtitle = "The nation-wide per-pupil spending was $18,460.", return = "plot")
```
:::

The same is true for spending per-pupil. @cr-spendingpp

::: {#cr-balance}
```{r}
state_map_var_discrete(us_map_data, account_balance, plotly_text, "Year Balance",
                       subtitle = "There was 21 billion dollars remaining at the end of school year.", return = "plot")
```
:::

There was **21 Billion** excess in revenue which was no spent. @cr-balance

::::::

::::::{.cr-section layout="overlay-center"}

:::{#cr-us}
```{r}
ggplot(us_map_data) +
  geom_sf(fill = "skyblue", color = "white") +
  #geom_sf_text(aes(label = iso3166_2), size = 2) +
  theme_void() +
  theme(
    plot.background = element_rect(fill =background_color,color = "transparent"),
      panel.background = element_rect(fill =background_color,color = "transparent"))
```
:::

So, why study **School Finance?** @cr-us

Because, school finance is an **Intervention** for creating better schools. @cr-us

And, a **lot of money** goes into US Education systems. @cr-us


**State policy** purposefully divides money among districts serving diverse student needs. @cr-us

Existing research has described the variation in all aspects of school funding.

However, it hasn't documents the variation in revenue allocation and spending for **Special Population, such as students in special education or English language programs** @cr-us


:::{#cr-or}


```{r}
or_school_district <- st_read(
  "data_raw/EDGE_SCHOOLDISTRICT_TL24_SY2324/EDGE_SCHOOLDISTRICT_TL24_SY2324.shp",
  quiet = TRUE) %>%
  janitor::clean_names() %>%
  filter(statefp == 41) %>%
  as.data.frame() %>%
  select(name, lograde, higrade, intptlon, intptlat) %>%
  mutate(across(c(intptlon, intptlat), parse_number),
         hover_text = paste0(name, "\n", "Grades: ", lograde,"-",higrade))

or_counties <- counties(cb = TRUE,
                        resolution = "20m",
                        class = "sf",
                        state = "Oregon",
                        progress_bar= FALSE) %>%
  janitor::clean_names()
```

```{r}
ggplot() +
  geom_sf(data = or_counties, fill = "#FEE123", color = "gray80", alpha = 0.5) +  # Districts
  theme_void() +
  theme(
      plot.background = element_rect(fill =background_color,color = "transparent"),
      panel.background = element_rect(fill =background_color,color = "transparent"))
```
:::

So, why study **Oregon?** @cr-or

Between **50-60%** Oregon's funding comes from state revenues. @cr-or

Oregon allocates revenue based on student subgroups

  - In programs, like Special education and English-language programs

  - With certain characteristics, like student in poverty, students in foster care/neglected or delinquent, students in parenting programs. @cr-or

And, Oregon tracks school-level spending based on student subgroups - After the 2017 state legislation. @cr-or

::::::

## **Explore the 197 Oregon School Districts**

```{r}
map <- ggplot() +
  geom_sf(data = or_counties, fill = "#FEE123", color = "gray80", alpha = 0.5) +  # Districts
  geom_point(data = or_school_district,
             aes(x = intptlon, y = intptlat, text = name), color = "#154733", size = 1) +
  theme_void() +
  theme(
    text = element_text(family = "serif"),
    axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Location of Oregon School Districts")

  ggplotly(map, tooltip = "name") %>%
    layout(showlegend = FALSE, paper_bgcolor = "rgba(0,0,0,0)", plot_bgcolor = "rgba(0,0,0,0)")
```
