---
title: "An Overview of Oregon's State Funding Allocation"
subtitle: "Subtitle"
editor: visual
author: "Havisha Khurana"
date: "today"
date-format: "full"
format: 
    html:
        embed-resources: true
        code-fold: true
        code-link: true
        theme: journal
        highlight: tango
        df-print: paged
        self-contained-math: true
        abstract: You can add an abstract here
        abstract-title: Abstract Title
        toc: true
        toc_float: true
        toc_location: "left"
        toc_expand: true
        code-overflow: wrap
        code-line-numbers: true
        code-copy: true
        footnotes-hover: true
        highlight-style: pygments
execute:
   warning: false
   error: false
   message: false
---

The state of Oregon uses a student-based funding model to allocate revenues to its 197 school districts. In this analysis, I will describe trends in its funding allocation from 2011 to 2024.

```{r}
# load libraries
library(tidyverse)
library(gghighlight)
library(geomtextpath)
library(patchwork)
library(modelsummary)
```

```{r}
district_year_adm <- readr::read_csv(here::here("data_clean/district_year_admw.csv")) %>%
  select(-1) %>%
  # make columns for iep related information
  #group_by(school_year, entry_name, district_name) 
  rowwise() %>%
  mutate(district_iep_students = 
           ifelse(grepl("IEP Students", category), parse_number(category),NA)) %>%
  ungroup() %>%
  group_by(district_id, school_year) %>%
  mutate(district_iep_students = sum(district_iep_students, na.rm = T),
         district_iep_percent = district_iep_students/current_year_adm_district, 
         district_iep_11 = sum(current_year_adm[grepl("IEP Students", category)]),
         district_iep_above_11 = sum(current_year_adm[grepl("IEP Above", category)]),
         district_iep_no_extra = district_iep_students - district_iep_11 - district_iep_above_11,
         district_iep_eff_weight = (district_iep_11 + district_iep_above_11)/district_iep_students
         )

hcd_year <-  readr::read_csv(here::here("data_clean/hcd_year_good.csv")) %>%
  select(-1)

hcd_all_year <- readr::read_csv("data_clean/hcd_state_2003_2022.csv") %>%
  select(-1)

district_details <- district_year_adm %>%
  distinct(school_year, district_id, district_name, current_year_adm_district)
```

```{r}
one_year_scatterplot <- function(df, year = 2011, 
                                 var_y, var_x, log_x_on = T, log_y_on = T,
                                lab_x = "", lab_y = "") {
  
  plot <- df %>%
  filter(school_year == year) %>%
  distinct(school_year, {{var_x}}, {{var_y}}) %>%
  mutate(across(-school_year), ~.x - mean(,x)) %>%
  ggplot(aes(x = {{var_x}}, y = {{var_y}})) +
  geom_smooth()+
  geom_vline(xintercept = 0) +
  geom_xline(y_intercept = 0) +
  geom_point() +
  labs(
    x = lab_x,
    y = lab_y
  ) 
  
  if(log_x_on == TRUE){
    plot <- plot +
      scale_x_continuous(trans = "log10",
                     breaks = c(1, 10, 100, 1000, 10000))
  }
  
  if(log_y_on == TRUE){
    plot <- plot +
      scale_y_continuous(trans = "log10",
                     breaks = c(1, 10, 100, 1000, 10000))
  }
  
  return(plot)
  
}
```

```{r}
plots_theme <- function(ggplot_object){
  ggplot_object <- ggplot_object +
  theme_minimal()  +
  theme(
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "azure", color = "transparent")
  ) +
    guides(color = "none",
           fill = "none")
  
  return(ggplot_object)
}
```

```{r}
cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = F)+
  gghighlight(max(current_year_adm_district) > 30000,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.3)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "District Student Membership Across Years")

cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = F)+
  gghighlight(max(current_year_adm_district) > 13000 & max(current_year_adm_district) < 30000,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.3)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "District Student Membership Across Years")

cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = T)+
  gghighlight(max(current_year_adm_district) < 10,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.3)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Log District Student Membership Across Years")
```


<!-- ```{r} -->
<!-- district_2024_adm_arrange <- district_year_adm %>% -->
<!--   filter(school_year == "2024") %>% -->
<!--   distinct(district_name, current_year_adm_district) %>% -->
<!--   arrange(desc(current_year_adm_district)) %>% -->
<!--   mutate(text = paste0(district_name,": ",round(current_year_adm_district))) -->
<!-- ``` -->

<!-- **District-ADM by school year**: District ADM exhibits similar distribution by academic year. In 2017, 2020 and 2021, there were districts with less than 1 student. The vast majority of districts have around 1000 students. The median number of students in a school district across years is close to 800 and the mean number of students in a school district across years is 2800. -->

<!-- The 10 biggest school districts in 2024 in Oregon were: `r head(district_2024_adm_arrange, 10) %>% pull(text)`. -->

<!-- The 10 smallest schools districts in 2024 in Oregon were: `r tail(district_2024_adm_arrange, 10) %>% pull(text)`. -->


```{r}
district_year_subgroup_enrollment <- district_year_adm %>%
  distinct(school_year, district_id, district_name, district_iep_students) %>%
  mutate(category = "IEP") %>%
  rename("current_year_adm" = "district_iep_students") %>%
  rbind(
  district_year_adm %>%
  filter(!grepl("IEP|School|Scholars", category)) %>%
  group_by(school_year, category, district_id, district_name) %>%
  summarize(current_year_adm = sum(current_year_adm))) %>%
  rbind(hcd_year %>%
          select(school_year, actual_hcd_count, dist_id) %>%
          rename("district_id" = "dist_id",
                 "current_year_adm" = "actual_hcd_count") %>%
          mutate(category = "HCD")) %>%
  mutate(category = gsub("Students in ", "", category)) %>%
  group_by(school_year, district_id) %>%
  mutate(
    prop_current_year_adm = current_year_adm/current_year_adm[category == "ADMr"],
    total = current_year_adm[category == "ADMr"]
  )
```


```{r}
state_year_subgroup_enrollment <- district_year_subgroup_enrollment %>%
  ungroup() %>%
  group_by(school_year, category) %>%
  summarize(total = sum(current_year_adm)) %>%
  rbind(hcd_all_year %>%
          filter(school_year >= 2011, school_year < 2018) %>%
          select(school_year, eligible_students) %>%
          rename("total" = "eligible_students") %>%
          mutate(category = "HCD")) %>%
  group_by(school_year) %>%
  mutate(
    percentage = round(total*100/total[category == "ADMr"],3)
  ) %>%
  filter(category != "ADMr", !is.na(percentage))
```
```{r}
ggplot(state_year_subgroup_enrollment, aes(x = school_year,
               node = category,
               fill = category,
               value = percentage)) +
  geom_sankey_bump(space = 0, type = "alluvial", color = "transparent", smooth = 6) +
  scale_fill_viridis_d(option = "A", alpha = .8) +
  theme_sankey_bump(base_size = 16) +
  labs(x = NULL,
       y = "Percentage",
       fill = NULL,
       color = NULL) +
  theme(legend.position = "bottom") +
  labs(title = "State-level subgroup representation")
```

```{r}
categories <- unique(district_year_subgroup_enrollment$category)[c(1,3:5)]

map(categories, ~cont_across_years_w_mean(df = filter(district_year_subgroup_enrollment, category == .x), var_x = school_year, var_y = prop_current_year_adm, log_on = F,
                                          mean_numerator = current_year_adm,
                                          mean_denominator = total) +
      labs(title = .x))
```

Proportion of subgroups by total enrollment <Quantile of enrollment and quantile of proportion by a category> Two ways to represent

```{r}
one_year_scatterplot_w_bars(df = district_year_subgroup_enrollment %>%
  filter(category == "ESL programs"),
   var_y = prop_current_year_adm, 
  var_x = total, log_x_on = T, log_y_on = F,
                                  add_bars = T,
                                 var_x_name = "District Membership", var_y_name = "Proportion of students in ESL programs"
)

```


```{r}
district_year_adm_summary <- district_year_adm %>%
  distinct(school_year, district_id, current_year_adm_district, current_year_admw_district, current_year_eff_admw_district) %>%
  mutate(
    adm_admw_ratio = current_year_admw_district/current_year_adm_district,
    eff_adm_admw_ratio = current_year_eff_admw_district/current_year_adm_district
  ) 

one_year_scatterplot_w_bars(df =district_year_adm_summary,
   var_y = current_year_admw_district, 
  var_x = current_year_adm_district, log_x_on = T, log_y_on = T,
                                  add_bars = F,
                                 var_x_name = "District Membership", var_y_name = "Proportion of students in ESL programs"
)

# library(lme4)
# 
# lmer(current_year_adm_district ~ 1 | district_id,
#      data = district_year_adm_summary)
#99.7% variance in the ADM is explained by the between-districts difference.
# The enrollment remains pretty stable between districts across time.
```

**What proportion of District-ADM do districts get?** The allocation rate (ADM-Weighted/ADM) per student has an exponential decay shape: Small districts (with less than 100 students) get between 2 to 9 times the allocation per student. Districts between 100 to 1000 students get between 1 to 2 times the allocation per student. The relationship is close to the 1 line for schools with higher than 1000 students. The median allocation ratio is `r median(district_year_adm_ratio$adm_admw_ratio)`. About 20% of the variation in the allocation ratio is explained at the between-district levels.

**How much do schools really get?** Since Oregon has a safeguard rule where the districts get max of their previous or current year weighted ADM, most school districts get funds allocated more than their resident schools in a given year. In the whole sample, only \<\> proportion of the times did districts get funds based on their current year ADM.

**What proportion of district ADM are categories that get exclusive funding?** **And variation that is explained by between district factors?**

Next set of plots:

-   Trends in Revenue: money from state, local, and federal sources - and variation.

```{r}
revenue <- readr::read_csv("data_clean/district_revenue_sources_2019_2022.csv") %>%
  select(-1) %>%
  left_join(district_details)

# dot-plot
revenue %>%
  filter(source == "federal") %>%
  mutate(ppe = revenue_dollar/current_year_adm_district) %>%
  cont_across_years_w_mean(df = ., var_y = ppe, var_x = school_year, log_on = T, mean_type = 2, mean_numerator = "", mean_denominator = "", add_dollar = T) 

cont_across_years_w_mean(df = revenue, var_y = ppe, var_x = school_year, log_on = T, mean_type = 2, mean_numerator = "", mean_denominator = "", add_dollar = T) 
 
revenue %>%
    distinct(school_year, district_id, district_name, ppe) %>%
    ggplot(aes(x = school_year, y = ppe)) +
    geom_line(aes(group = district_name), color = "gray80", alpha = 0.3) +
    #scale_y_log10()+
    # geom_smooth(se = F)+
    # scale_y_log10(
    #   breaks = c(1,10,100,1000, 10000, 50000)
    # ) +
    theme_minimal()+
    theme(
      panel.grid = element_blank(),
      axis.ticks = element_line()
    ) +
    guides(color = "none") +

# facet-plot,
ggplot(revenue, aes(y = per_revenue, x = source, color = source)) +
  geom_jitter(alpha = 0.5) +
  facet_wrap(~school_year) +
  scale_y_continuous(limits = c(0,1))

revenue %>%
  #filter(source != "other") %>%
  group_by(school_year, source) %>%
  summarize(state_total = sum(revenue_dollar)) %>%
  mutate(prop = state_total/sum(state_total)) %>%
  ggplot(aes(y = factor(school_year), x = prop, fill = source)) +
  geom_col()

lmer(total_revenue ~ 1 | district_id,
     distinct(revenue, school_year, district_id, total_revenue))

lmer(per_revenue ~ 1 | district_id,
     distinct(filter(revenue, source == "intermediate"), school_year, district_id, per_revenue))

# most variation in revenue is explained at the between district levels, like 98% - local and state more stable. federal, other, intermediate more variable.
```

-   Trends in aggregated expenditure by categories defined in the dataset - and variation

```{r}
#df ready
```

Main interest: funding for and expenditure on two groups: SpEd and EL

-   funding: HCD

```{r}
ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = threshold)) +
  geom_col()

ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = grant_amount)) +
  geom_col()

ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = eligible_students)) +
  geom_col() #percent of total, already shown. between 0.5 to 1% from 2011 to 2022

ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = total_expenditre)) +
  geom_col()

# total expenditure per student
ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = total_expenditre/eligible_students)) +
  geom_col()

# total expenditure per student
ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = expenditure_above_threshold/eligible_students)) +
  geom_col()

#
ggplot(hcd_all_year,
       aes(y = factor(school_year),
           x = calculated_rate)) +
  geom_col()
```


-   prop of iep adm that's from 11% and that's on top

```{r}

```


-   what proportion of oregon students get funding through the double funding route?

-   Expenditure: Use same criteria to make the aggregated categories but separate for flagged and not, and

-   Evaluate discontinuity
