---
title: "An Overview of Oregon's State Funding Allocation"
format: closeread-html
---

The state of Oregon uses a student-based funding model to allocate revenues to its 197 school districts. In this article, I will describe trends in funding allocation from 2011 to 2024. 


```{r}
# load libraries
library(tidyverse)
library(plotly)
```


```{r}
district_year_adm <- readr::read_csv(here::here("data_clean/district_year_admw.csv")) %>%
  select(-1) %>%
  # make columns for iep related information
  #group_by(school_year, entry_name, district_name) 
  rowwise() %>%
  mutate(district_iep_students = 
           ifelse(grepl("IEP Students", category), parse_number(category),NA)) %>%
  ungroup() %>%
  group_by(district_id, school_year) %>%
  mutate(district_iep_students = sum(district_iep_students, na.rm = T),
         district_iep_percent = district_iep_students/current_year_adm_district, 
         district_iep_11 = sum(current_year_adm[grepl("IEP Students", category)]),
         district_iep_above_11 = sum(current_year_adm[grepl("IEP Above", category)]),
         district_iep_no_extra = district_iep_students - district_iep_11 - district_iep_above_11,
         district_iep_eff_weight = (district_iep_11 + district_iep_above_11)/district_iep_students
         )

hcd_year <-  readr::read_csv(here::here("data_clean/hcd_year_good.csv")) %>%
  select(-1)
```

```{r}
year_dotplot <- function(df, var_y , var_x, log_on = T,
                         lab_x = "", lab_y = "Year"){
  
  plot <- district_year_adm %>%
  distinct(district_id, {{var_y}}, {{var_x}}, district_name) %>%
  ggplot(aes(y = fct_rev(factor({{var_y}})), x = {{var_x}}, color = {{var_x}})) +
  geom_jitter(show.legend = F, alpha = 0.4) +
  geom_boxplot(outliers = F, fill = "transparent", show.legend = F)+
  labs(
    x = lab_x,
    y = lab_y
  ) +
  theme_minimal() +
 scale_color_continuous(type = "viridis")
  
  if(log_on == TRUE){
    plot <- plot +
      scale_color_continuous(trans = "log10", type = "viridis")+
      scale_x_continuous(trans = "log10",
                     breaks = c(1, 10, 100, 1000, 10000))
      
  }

  return(plot)
}

one_year_scatterplot <- function(df, year = 2011, 
                                 var_y, var_x, log_x_on = T, log_y_on = T,
                                lab_x = "", lab_y = "") {
  
  plot <- df %>%
  filter(school_year == year) %>%
  distinct(school_year, {{var_x}}, {{var_y}}) %>%
  ggplot(aes(x = {{var_x}}, y = {{var_y}})) +
  geom_point() +
  labs(
    x = lab_x,
    y = lab_y
  ) +
  theme_minimal()  +
  theme(
    panel.grid.minor = element_blank()
  )
  
  if(log_x_on == TRUE){
    plot <- plot +
      scale_x_continuous(trans = "log10",
                     breaks = c(1, 10, 100, 1000, 10000))
  }
  
  if(log_y_on == TRUE){
    plot <- plot +
      scale_y_continuous(trans = "log10",
                     breaks = c(1, 10, 100, 1000, 10000))
  }
  
  return(plot)
  
}
```

:::{.cr-section}

:::{#cr-stateprop}
```{r}

```
:::


District-ADM by school year

```{r}
year_dotplot(district_year_adm,
             var_y = school_year,
             var_x = current_year_adm_district,
             log_on = T,
             lab_x = "District Average Daily Membership (ADM)") 
```

What proportion of District-ADM do districts get?

```{r}
district_year_adm %>%
  distinct(school_year, district_id, current_year_adm_district, current_year_admw_district) %>%
  mutate(
    adm_admw_ratio = current_year_admw_district/current_year_adm_district
  ) %>%
  one_year_scatterplot(df = ., var_y = adm_admw_ratio, var_x = current_year_adm_district, log_y_on = F,
                       lab_x = "District ADM",
                       lab_y = "Allocation rate per student") +
  scale_y_continuous(limits = c(0,10),
                     breaks = 0:10)

district_year_adm %>%
  distinct(school_year, district_id, current_year_admw_district, current_year_eff_admw_district) %>%
  mutate(extra = current_year_eff_admw_district - current_year_admw_district) %>%
  one_year_scatterplot(df = ., var_y = extra, var_x = current_year_admw_district, log_y_on = F, log_x_on = T,
                       lab_x = "District Current-year ADMw",
                       lab_y = "District Allocated ADMw (max of this or previous yeaar)")
```

What proportion of district ADM are categories that get exclusive funding?

```{r}
district_year_subgroup_enrollment <- district_year_adm %>%
  distinct(school_year, district_id, district_name, district_iep_students) %>%
  mutate(category = "IEP") %>%
  rename("current_year_adm" = "district_iep_students") %>%
  rbind(
  district_year_adm %>%
  filter(!grepl("IEP|School|Scholars", category)) %>%
  group_by(school_year, category, district_id, district_name) %>%
  summarize(current_year_adm = sum(current_year_adm))) %>%
  rbind(hcd_year %>%
          select(school_year, actual_hcd_count, dist_id) %>%
          rename("district_id" = "dist_id",
                 "current_year_adm" = "actual_hcd_count") %>%
          mutate(category = "HCD")) %>%
  mutate(category = gsub("Students in ", "", category)) %>%
  group_by(school_year, district_id) %>%
  mutate(
    prop_current_year_adm = current_year_adm/current_year_adm[category == "ADMr"],
    total = current_year_adm[category == "ADMr"]
  )
```


```{r}
district_year_subgroup_enrollment %>%
  ungroup() %>%
  group_by(school_year, category) %>%
  summarize(total = sum(current_year_adm)) %>%
  group_by(school_year) %>%
  mutate(
    percentage = round(total*100/total[category == "ADMr"],3)
  ) %>%
  filter(category != "ADMr") %>%
  ggplot(data = ., aes(x = percentage,
             y = fct_reorder(factor(category), percentage))) +
  geom_col(fill = "cornflowerblue") +
  geom_text(aes(label = round(percentage,1)), nudge_x = 2, size = 2, color = "gray30") +
  facet_wrap(~school_year, ncol = 3) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_text(lineheight = 0.5),
    axis.text.x = element_blank()
  ) 
```

Proportion of subgroups by total enrollment

```{r}
categories <- unique(district_year_subgroup_enrollment$category)[-2]

map(categories,
    ~one_year_scatterplot(
      year = 2019,
      df = filter(district_year_subgroup_enrollment, category == .x),
      var_x = total,
      var_y = prop_current_year_adm,
      log_y_on = F,
      lab_x = .x))



map(2011:2023, ~one_year_scatterplot(year = .x,
                                     df = district_year_adm,
                     var_x = district_iep_percent,
                     var_y = district_iep_eff_weight,
                     log_y_on = F,
                     log_x_on = F))
```



```{r}

```

:::