---
title: "An Overview of Oregon's State Funding Allocation"
editor_options: 
  chunk_output_type: console
author: "Havisha Khurana"
date: "today"
date-format: "full"
format: 
    html:
        embed-resources: true
        theme: journal
        highlight: tango
        df-print: paged
        self-contained-math: true
        toc: true
        toc_float: true
        toc_location: "left"
        toc_expand: true
        code-overflow: wrap
        code-line-numbers: true
        code-copy: true
        footnotes-hover: true
        highlight-style: pygments
execute:
   warning: false
   error: false
   message: false
   echo: false
---

In Oregon, state and local revenues are pooled together and redistributed to school districts based on a funding formula. Approximately 60% of the total education budget comes from the state’s share, making it a critical revenue source for most districts. The **State School Fund (SSF)** allocates lump-sum funding to districts based on the number of resident students, adjusted for differential needs as specified by the Oregon Revised Statutes (ORS 327.008). Students with disabilities are double-weighted in the funding formula, with districts guaranteed double-weighted funding for up to 11% of their student population. Other student groups, such as students classified as English Language Learners or those qualifying for Free-or-reduced price lunch are also proportionately weighted in the formula. While state allocations are tied to student demographics, districts have the flexibility to use these revenues as they see fit. 

Another state fiscal policy to support students with disabilities is the **High-Cost Disability Fund**, which reimburses districts for expenses exceeding \$30,000 per special education student (the cap was at \$25,000 from 2003-2005). Each year, a portion of the SSF, amounting to less than 5% of the total appropriation, is designated for this fund. If statewide claims exceed the available funds, reimbursements are prorated across districts. 

In this analysis, I will describe descriptive trends in district enrollment, districts' state-funding allocation, district revenue, and district expenditure related to students with disabilities whose services cost more than a given threshold and are paid by the grant called *High-Cost Disability Fund*.

```{r}
# load libraries
library(tidyverse)
library(gghighlight)
#library(geomtextpath)
library(patchwork)
```

```{r}
# load r with plots code
source("code/writeup_plots.R")
```

## District ADM Data

*District-level allocations from the State School Fund (SSF), disaggregated by student type, covering the years 2011–2024.*

```{r}
# data prep
district_year_adm <- readr::read_csv(here::here("data_clean/district_year_admw.csv")) %>%
  select(-1) %>%
  # make columns for iep related information
  #group_by(school_year, entry_name, district_name) 
  rowwise() %>%
  mutate(district_iep_students = 
           ifelse(grepl("IEP Students", category), parse_number(category),NA)) %>%
  ungroup() %>%
  group_by(district_id, school_year) %>%
  mutate(district_iep_students = sum(district_iep_students, na.rm = T),
         district_iep_percent = district_iep_students/current_year_adm_district, 
         district_iep_11 = sum(current_year_adm[grepl("IEP Students", category)]),
         district_iep_above_11 = sum(current_year_adm[grepl("IEP Above", category)]),
         district_iep_no_extra = district_iep_students - district_iep_11 - district_iep_above_11,
         district_iep_eff_weight = (district_iep_11 + district_iep_above_11)/district_iep_students
         )

district_details <- district_year_adm %>%
  distinct(school_year, district_id, district_name, current_year_adm_district)


hcd_year <-  readr::read_csv(here::here("data_clean/hcd_year_good.csv")) %>%
  select(-1) %>%
  rename("district_id" = "dist_id") %>%
  left_join(district_details)

hcd_all_year <- readr::read_csv("data_clean/hcd_state_2003_2022.csv") %>%
  select(-1) 
```

### Trends in Enrollment 

```{r}
cont_across_years_w_mean(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             mean_type = 2,
             mean_numerator = "",
             mean_denominator = "",
             log_on = T,
             round_pos = 0,
             lab_y = "Log of Student Membership") +
  labs(
    title = "Oregon Enrollment Trends Across 14 years",
    subtitle = str_wrap("An average Oregon school districts’ enrollment is around 2800 students, with substantial variation around the mean", 70)
  )
```

An average Oregon school districts’ enrollment is around 2800 students- which remains pretty stable across the years. However, there is substantial variation in enrollment between districts. 


```{r}
cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = F,
             lab_y = "Student Membership")+
  gghighlight(max(current_year_adm_district) > 30000,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.5)) +
  scale_color_brewer(palette = "Dark2") +
   labs(
    title = "Oregon Enrollment Trends Across 14 years",
    subtitle = str_wrap("Portland, Salem-Keizer, and Beaverton are the biggest school districts with more than 35000 students.", 70)
  )
```

```{r}
cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = F,
             lab_y = "Student Membership")+
  gghighlight(max(current_year_adm_district) > 13000 & max(current_year_adm_district) < 30000,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.5)) +
  scale_color_brewer(palette = "Dark2")+
  labs(
    title = "Oregon Enrollment Trends Across 14 years",
    subtitle = str_wrap("A handful of districts have more than 13000 students, including Eugene SD 4J", 70)
  )
```


```{r}
cont_across_years(district_year_adm,
             var_x = school_year,
             var_y = current_year_adm_district,
             log_on = T,
             lab_y = "Log of Student Membership")+
  gghighlight(max(current_year_adm_district) < 10,
              label_key = district_name,
              unhighlighted_params = list(alpha = 0.5)) +
  scale_color_brewer(palette = "Dark2")+
  labs(
    title = "Oregon Enrollment Trends Across 14 years",
    subtitle = str_wrap("About 13% of OR school districts have less than 100 students.", 70)
  )
```


### State Trends for Subgroup Enrollment

```{r}
district_year_subgroup_enrollment <- district_year_adm %>%
  distinct(school_year, district_id, district_name, district_iep_students) %>%
  mutate(category = "IEP") %>%
  rename("current_year_adm" = "district_iep_students") %>%
  rbind(
  district_year_adm %>%
  filter(!grepl("IEP|School|Scholars", category)) %>%
  group_by(school_year, category, district_id, district_name) %>%
  summarize(current_year_adm = sum(current_year_adm))) %>%
  rbind(hcd_year %>%
          select(school_year, actual_hcd_count, district_id, district_name) %>%
          rename("current_year_adm" = "actual_hcd_count") %>%
          mutate(category = "HCD")) %>%
  mutate(category = gsub("Students in ", "", category)) %>%
  group_by(school_year, district_id) %>%
  mutate(
    prop_current_year_adm = current_year_adm/current_year_adm[category == "ADMr"],
    total = current_year_adm[category == "ADMr"]
  )
```

```{r}
state_year_subgroup_enrollment <- district_year_subgroup_enrollment %>%
  ungroup() %>%
  group_by(school_year, category) %>%
  summarize(total = sum(current_year_adm)) %>%
  rbind(hcd_all_year %>%
          filter(school_year >= 2011, school_year < 2018) %>%
          select(school_year, eligible_students) %>%
          rename("total" = "eligible_students") %>%
          mutate(category = "HCD")) %>%
  group_by(school_year) %>%
  mutate(
    percentage = round(total*100/total[category == "ADMr"],3)
  ) %>%
  filter(category != "ADMr", !is.na(percentage))
```

```{r}
ggplot(filter(state_year_subgroup_enrollment, school_year < 2023),
       aes(x = school_year,
               node = category,
               fill = category,
               value = percentage)) +
  geom_sankey_bump(space = 0, type = "alluvial", color = "transparent", smooth = 6,
                   alpha = 0.6) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = 2011:2022) +
  theme_sankey_bump() +
  annotate("label",
           x = 2012,
           label = "Poverty",
           y = 35)+
    annotate("label",
           x = 2012,
           label = "IEP",
           y = 19)+
    annotate("label",
           x = 2012,
           label = "EL",
           y = 8)+
  labs(x = NULL,
       y = "Percentage",
       fill = NULL,
       color = NULL,
       title = "State-level subgroup representation",
       subtitle = str_wrap("Around 40% of Oregon’s students are in programs or demographic categories that get additional funding weights from the state-funding formula.", 60)) +
  theme(legend.position = "bottom") +
  guides(color = "none",
         fill = "none")
```

The three biggest groups are students under poverty, students served in special education and in English language programs.

In 2011, about 16% of the all students were under poverty, 14% of all students were in special education programs and 9% of students were in the ESL programs. In 2024, students served in special education was the biggest group at 15% followed by students under poverty and students in ELP at around 12%. About 1% of the students are in other programs. 

### District Trends for Subgroup Enrollment

Next, we look at the trends percentage of subgroup representation across districts over time. 

```{r}
categories <- unique(district_year_subgroup_enrollment$category)[c(1,3:7)]

pwalk(list(categories, 
      c("Special Education (or with an Individualized Education Plan aka IEP)",
        "English as a Second Language Program",
        categories[3:5],
        "Students Eligible for High Cost Disability Fund"),
       1:length(categories)*5),
      ~print(cont_across_years_w_mean(df = filter(district_year_subgroup_enrollment, category == ..1), var_x = school_year, var_y = prop_current_year_adm, log_on = F,
                                          mean_numerator = current_year_adm,
                                          mean_denominator = total,
             lab_y = "Proportion of District Membeship", seed = ..3) +
      labs(title = "District-level subgroup representation",
            subtitle = ..2)))
```

Students in special education (IEP or students with an Individualized Education Plans) and English language programs  have been steadily growing over the years, and those classified in poverty are steadily decreasing, but there is plenty of between-district variation. 

Between 0 to 3% of students were classified as HCD (eligible for High Cost Disability Fund) between districts between 2018 to 2022.

### Relationship between Districts Total Enrollment and Subgroup Enrollment

```{r}

walk2(categories, 1:length(categories)*5, ~one_year_scatterplot_w_bars(df = filter(district_year_subgroup_enrollment, category == .x), var_x = total, var_y = prop_current_year_adm, log_x_on = T, 
                                          log_y_on = F,
                                          var_x_name = "District Membership", 
                                          var_y_name = .x,
                                  add_bars = T, seed = .y)%>% print())

```

For students with an IEP and student in poverty, there is variation in proportion of the subgroup relative to the school enrollment. However, the average sub-group representation by quintiles of school enrollment appears even, at between 12-16%. A handful of school districts with an enrollment of around 10 students served no special education students in their districts in 2022. Some of the low-enrollment schools have high representation of students in poverty.

Students in ESL programs, and those in foster care or neglected/delinquent, pregnant and parenting programs, and those eligible for reimbursements from the high-cost disability funds are more concentrated in bigger schools.

## District Funds Allocation Data

I define **allocation ratio** as the ratio of weighted average daily membership to the real average daily membership. Since students with some characteristics are weighted more than 1 per the state formula, a district gets resources as if there were more students in the school. So, an allocation ratio of 2 means that districts get 2 student equivalent for every 1 real student in their district.

```{r}
district_year_adm_summary <- district_year_adm %>%
  distinct(school_year, district_id, current_year_adm_district, current_year_admw_district, current_year_eff_admw_district) %>%
  mutate(
    adm_admw_ratio = current_year_admw_district/current_year_adm_district,
    eff_adm_admw_ratio = current_year_eff_admw_district/current_year_adm_district,
    difference_ratio = eff_adm_admw_ratio - adm_admw_ratio
  ) 

one_year_scatterplot_w_bars(df = district_year_adm_summary, 
                            var_x = current_year_adm_district, 
                            var_y = adm_admw_ratio, 
                            log_x_on = T, 
                            log_y_on = F,
                            var_x_name = "Current Year District Membership", 
                            var_y_name = "Allocation Ratio",
                                  add_bars = T,
                            seed = 3)


# one_year_scatterplot_w_bars(df = district_year_adm_summary, 
#                             var_x = current_year_adm_district, 
#                             var_y = eff_adm_admw_ratio, 
#                             log_x_on = T, 
#                             log_y_on = F,
#                             var_x_name = "Current Year District Membership", 
#                             var_y_name = "Effective Allocation Ratio",
                                  # add_bars = T)

# one_year_scatterplot_w_bars(df = district_year_adm_summary, 
#                                         
#                             var_x =current_year_adm_district, 
#                             var_y =  difference_ratio, 
#                             log_x_on = T, 
#                             log_y_on = F,
#                             var_x_name = "Current Year Enrollment", 
#                             var_y_name = "Difference: Effective and Calculated Allocation Ratio",
#                                   add_bars = T)

# library(lme4)
# 
# lmer(current_year_adm_district ~ 1 | district_id,
#      data = district_year_adm_summary)
#99.7% variance in the ADM is explained by the between-districts difference.
# The enrollment remains pretty stable between districts across time.
```

Students in smaller school districts get more per-student revenue allocated (in student units) than students in larger districts. In the first enrollment quintile, school districts get about 4 student equivalent for each of their student. In the second quintile, school districts get about 2 student equivalent for each of their student. In larger districts, school districts get 1.3 student equivalent for each of their student. 

```{r}
#| eval: false
# **What proportion of District-ADM do districts get?** The allocation rate (ADM-Weighted/ADM) per student has an exponential decay shape: Small districts (with less than 100 students) get between 2 to 9 times the allocation per student. Districts between 100 to 1000 students get between 1 to 2 times the allocation per student. The relationship is close to the 1 line for schools with higher than 1000 students. The median allocation ratio is `r median(district_year_adm_ratio$adm_admw_ratio)`. About 20% of the variation in the allocation ratio is explained at the between-district levels.
# 
# **How much do schools really get?** Since Oregon has a safeguard rule where the districts get max of their previous or current year weighted ADM, most school districts get funds allocated more than their resident schools in a given year. In the whole sample, only proportion of the times did districts get funds based on their current year ADM.
# 
# **What proportion of district ADM are categories that get exclusive funding?** **And variation that is explained by between district factors?**

```

## Revenue Data

```{r}
revenue <- readr::read_csv("data_clean/district_revenue_sources_2019_2022.csv") %>%
  select(-1) %>%
  left_join(district_details) %>%
  mutate(source = factor(source, 
                         levels = c("federal", "state", "local", "intermediate", "other")))
```

### State Trends in Per-pupil revenue

```{r}
#how much money?
revenue %>%
  group_by(school_year) %>%
  summarize(total_year = sum(revenue_dollar, na.rm = T)) %>%
  ggplot(., aes(x = school_year, y = total_year)) +
  geom_col(alpha = 0.7, fill = "#A6761D") +
  scale_y_continuous(
    breaks = c(0,5,10,15,20)*1000000000,
    labels = c("0", "5B", "10B", "15B", "20B")
    ) +
  labs(x = NULL,
       y = "Total Revenue",
       title = "Total Revenue Over Years",
       subtitle = str_wrap("17-20 Billion Dollars revenue was generated by school districts, although the amount could be biased upward due to COVID-19 relief in these years.", 60)) +
  theme_minimal()
```

```{r}
# state-revenue trends
revenue %>%
  group_by(school_year, source) %>%
  summarize(
    total_source = sum(revenue_dollar, na.rm = T)
  ) %>%
  group_by(school_year) %>%
  mutate(total_year = sum(total_source),
         per_category = total_source/total_year) %>%
  ggplot(., aes(x = school_year, y = per_category, fill = source)) +
  geom_col(alpha = 0.7) +
  geom_text(aes(label = round(per_category, 2),
                color = per_category > 0.1),
            position = position_fill()) +
  scale_fill_brewer(palette = "Dark2",
                    labels = str_to_title(levels(revenue$source))) +
  scale_color_manual(values = c("transparent","gray10")) +
  guides(
    color = "none"
  ) +
  labs(
    x = "",
    y = "Proportion of Revenue",
    fill = NULL,
    title = "Disaggregated Revenue by Source",
    subtitle = str_wrap("State, local, and other sources contribute a larger proportion towards overall revenue. Other sources mostly constitute of unused-balance from prior years.", 60)
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    legend.position = "top"
  )
```

```{r}
# without other
revenue %>%
  filter(source != "other") %>%
  group_by(school_year, source) %>%
  summarize(
    total_source = sum(revenue_dollar, na.rm = T)
  ) %>%
  group_by(school_year) %>%
  mutate(total_year = sum(total_source),
         per_category = total_source/total_year) %>%
  ggplot(., aes(x = school_year, y = per_category, fill = source)) +
  geom_col(alpha = 0.7) +
  geom_text(aes(label = round(per_category, 2),
                color = per_category > 0.05),
            position = position_fill()) +
  scale_fill_brewer(palette = "Dark2",
                    labels = str_to_title(levels(revenue$source))) +
  scale_color_manual(values = c("transparent","gray10")) +
  guides(
    color = "none"
  ) +
  labs(
    x = "",
    y = "Proportion of Revenue",
  fill = NULL,
      title = "Disaggregated Revenue by Source",
    subtitle = str_wrap("Considering only new revenue, state sources contribute to over half of the overall revenue", 60)
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    legend.position = "top"
  )
```

School districts had an average total revenue of 18 billion dollars for the four years I had publically-available data. Of the total revenue, between 28-33% came from State resources, 23-25% came from local resources, and 35-40% came from Other resources. Intermediate and federal resources accounted for the remaining. 

Almost all of state revenue is allocated using the SSF. Local revenue could be tax levies etc. The biggest contributor to Other resources are districts’ account balance from previous years – thus, districts saved substantial money from past years. The decrease in other sourcecs over the 4 years could be related to using emergency resources in response to COVID-19 related disruptions. 

If we remove the other sources and only consider the new money districts get, about 50% comes from state resources and 40 percent come from local resources. Between 6 to 10% percent come from federal sources. 

### Trends in Per-pupil revenue by Districts based on sources

```{r}
revenue_sources <- unique(revenue$source)

walk2(revenue_sources,
      1:length(revenue_sources)*5,
      ~print(cont_across_years_w_mean(df = filter(revenue, source == .x) %>%
                                                 mutate(ppe = revenue_dollar/current_year_adm_district), 
                                               var_x = school_year, 
                                               var_y = ppe, log_on = T,
                                               mean_type = 2,
                                          mean_numerator = "",
                                          mean_denominator = "",
                                          add_dollar = T,
                                          lab_y = "Per-pupil revenue",
                                          round_pos = 0,
                                seed = .y) +
      labs(title = paste0("From ", str_to_title(.x)," sources"))))
```

The average per-pupul revenue from all sources, except intermediate, went up from 2019 to 2022. Although the most dramatic increase in revenue was from federal sources, which went up about three times between 2019 to 2021. Revenue from local and intermediate sources remained more consistent over time.

### Trends in Revenue by District Enrollment

```{r}
walk2(revenue_sources,
      1:length(revenue_sources)*5, ~one_year_scatterplot_w_bars(df = filter(revenue, source == .x) %>%
                                                 mutate(ppe = revenue_dollar/current_year_adm_district), 
                                               var_x = current_year_adm_district, 
                                               var_y = ppe, 
                                               log_x_on = T,
                                               log_y_on = F,
                                              var_x_name = "Current Year Enrollment", 
                            var_y_name = paste0("Per-pupil Revenue: ",str_to_title(.x)),
                                  add_bars = T, seed = .y, round_dig = 0) %>% print())
```


Districts with smaller enrollment have a much higher per-pupil revenue from all sources compared to big districts. The most dramatic difference is in per-pupil revenue by other sources.

## HCD Population

### Students eligible for HCD and Grant Amount

```{r}
hcd_eligible <- hcd_bar_plot(df = hcd_all_year,
             var_x = eligible_students,
             var_label = eligible_students,
             title = "Eligible Students",
             nudge = -300, seed = 2)

hcd_grant <- hcd_bar_plot(df = mutate(hcd_all_year, grant_amount_text = gsub("0{1,}","M", grant_amount)),
             var_x = grant_amount,
             var_label = grant_amount_text,
             title = "Grant Amount", seed = 2,
             position = "right")


hcd_eligible + hcd_grant
```

The number of students identified as eligible for 'High Cost Disability Fund' have doubled from 2003 to 2022, or from 0.5% of the enrollment in 2003 to 1% in 2022. The grant money has also increased in phases. The drop in the number of students eligible for HCD in 2005 is because the eligibility threshold for expenditure increased from 25k to 30k in 2005. 

### Students eligible for HCD and Per-pupil expenditure above threshold

```{r}
expense_per_student <- hcd_bar_plot(df = mutate(hcd_all_year, ppe = expenditure_above_threshold/eligible_students),
             var_x = ppe,
             var_label = scales::dollar(round(ppe, -2)),
             title = "Average Expense per student", seed = 2,
             position = "right", nudge = -1000)

hcd_eligible + expense_per_student 
```

Per-pupil expenditure over the threshold was 9000 in 2003 and was 25000 in 2022.

### Per-pupil reimbursment and Per-pupil expenditure above threshold

```{r}
reimbusement_per_student <- hcd_bar_plot(df = mutate(hcd_all_year, ppe = expenditure_above_threshold*calculated_rate/eligible_students),
             var_x = ppe,
             var_label = scales::dollar(round(ppe, -2)),
             title = "Average Reimbursement per student", seed = 2,
             position = "left", nudge = -1000)

reimbusement_per_student + expense_per_student
```

Reimbursement hasn’t increases as much as the expenses. The gap in per-student expenditure vs reimbursement was $3k in 2003 and it was 15k in 2022. 
 
```{r}
#| eval: false

# -   Trends in aggregated expenditure by categories defined in the dataset - and variation
# 
# Main interest: funding for and expenditure on two groups: SpEd and EL
# 
# 
# <Is the 11% clean variation, or 2 year max a clean variation? If they are gaming one, are they gaming the other?>
# The district level optimization around the rules.
# 
# -   prop of iep adm that's from 11% and that's on top
# 
# -   what proportion of oregon students get funding through the double funding route?
# 
# -   Expenditure: Use same criteria to make the aggregated categories but separate for flagged and not, and
# 
# -   Evaluate discontinuity
```
